---
# ======== DOCUMENT METADATA ========
title: "<strong>Quarto Template for the Automated Diagnostic Analysis of Tabular Datasets</strong>" # Document title
subtitle: "<strong><em>Report for Exploratory Data Analysis, Data Quality Assessment, and Descriptive Summary</em></strong>" # Document subtitle
author:
  - "Full Name<br><span style='margin-left:1em'>Student ID (optional)<br><span style='margin-left:1em'>Email Address (optional)" # Author 1
  - "Full Name<br><span style='margin-left:1em'>Student ID (optional)<br><span style='margin-left:1em'>Email Address (optional)" # Author 2
abstract: |
  This Quarto template provides a reproducible workflow for the automated diagnostic analysis of tabular datasets. It generates a comprehensive exploratory report including data import, data quality assessment, descriptive statistics, missing-value analysis, variable distribution summaries, and association analyses. The template is designed to support rapid preliminary exploration of datasets while promoting transparency, reproducibility, and consistency across analyses. Its parameterized structure allows users to apply the same workflow to different datasets with minimal modifications.
lang: en             # HTML language
date: today          # Generation date
date-format: long    # Long date format

# ======== DOCUMENT PARAMETERS ========
params:
  data_path: "BIG5_data/BIG5/data.csv"   # Dataset to analyse. Relative paths are resolved from the project root; absolute paths are used as-is.
  data_type: "csv"                       # Supported file types: "csv" (including TSV and other delimited text files) or "excel"
  data_delim: "\t"                       # Column separator used only when data_type = "csv" (e.g. ",", ";" or "\t")
  data_sheet: 1                          # Worksheet name or index used only when data_type = "excel"

# ======== HTML OUTPUT SETTINGS ========
format:
  html:
    polyfill: false           # Disable external JavaScript polyfills
    embed-resources: true     # Embed all CSS/JS resources directly into the HTML file
    standalone: true          # Generate a self-contained HTML document
    toc: true                 # Enable the Table of Contents
    number-sections: true     # Automatically number section headings
    toc-depth: 5              # Include headings up to level 5 in the TOC
    code-tools: true          # Add the "Show Source" button for code chunks
    code-fold: true           # Make code chunks collapsible
    fig-cap: true             # Enable figure captions below plots and images

# ======== CHUNK EXECUTION OPTIONS ========
execute:
  echo: true                  # Display R code in the HTML output
  warning: false              # Hide warning messages during execution
  message: false              # Hide package startup and informational messages

# ======== CUSTOM CSS ========
include-in-header:
  - text: |
      <style>

      /* ---------- GLOBAL DOCUMENT STYLE ----------
         Base typography and general appearance.
         Defines font family, spacing, colours,
         links and default visual consistency.
         ------------------------------------------ */
      body{
        font-family:Georgia,"Times New Roman",serif; /* Academic serif font */
        line-height:1.6;                             /* Improve text readability */
        color:#222;                                  /* Soft black text colour */
        margin:50px 0 0;}                            /* Top spacing below header */

      /* ---------- LINKS ----------
         Consistent hyperlink colour scheme.
         ------------------------------------------ */
      a,
      .toc a{
        color:#004a99;              /* Professional blue links */
        text-decoration:none;}      /* Remove default underline */
      a:hover,
      .toc a:hover{
        color:#002b7f;              /* Darker blue on hover */
        text-decoration:underline;} /* Visual feedback */

      /* ---------- GLOBAL ROUNDED CORNERS ----------
         Apply a consistent rounded style to
         interface elements and containers.
         ------------------------------------------ */
      *{border-radius:5px;}
      /* Avoid rounded corners on textual elements
         to preserve an academic document style. */
      p,
      h1,h2,h3,h4,h5,h6{border-radius:0;}

      /* ---------- PARAGRAPH ALIGNMENT ----------
         Justified text for a report-like layout.
         ------------------------------------------ */
      p{text-align:justify;}


      /* ---------- CONTENT WIDTH ----------
         Controls the maximum width of the main
         document area for better readability.
         ------------------------------------------ */
      main{max-width:900px;} /* Limit text width */

      /* ---------- TABLE OF CONTENTS ----------
         Professional navigation box styling.
         Improves visibility and hierarchy of
         the document structure.
         ------------------------------------------ */
      #toc,
      #contents{
        position:relative;                     /* Keep element in normal flow */
        display:block;                         /* Standard block layout */
        width:90%;                             /* Leave margins around the box */
        margin:20px auto 40px;                 /* Vertical spacing */
        padding:10px;                          /* Internal spacing */
        background:#fafbfc;                    /* Neutral background */
        border:1px solid #d8dee9;              /* Light grey border */
        border-left:5px solid #002b7f;         /* Main colour accent */
        box-shadow:0 2px 6px rgba(0,0,0,.05);} /* Subtle depth effect */
      #TOC{z-index:1;}                         /* Keep TOC above overlapping elements */
      .toc{
        padding:10px;                          /* Inner spacing */
        background:transparent;}               /* Avoid duplicated background */
      .toc ul{list-style:none;}                /* Remove default bullets */
      .toc li{
        border-left:2px solid #d9d9d9;         /* Visual hierarchy indicator */
        padding-left:8px;}                     /* Distance from separator */

      /* ---------- CODE ----------
         Styling for source code, console output
         and computational results.
         Preserves formatting and horizontal
         scrolling for long lines.
         ------------------------------------------ */
      pre,
      pre code,
      code.sourceCode,
      .cell-output,
      .cell-output pre,
      .cell-output code{
        font-family:monospace;             /* Code-specific font */
        display:block;                     /* Preserve block formatting */
        white-space:pre!important;         /* Keep original spacing */
        word-break:keep-all!important;     /* Avoid breaking code lines */
        overflow-x:auto;                   /* Allow horizontal scrolling */
        max-width:100%;}                   /* Prevent container overflow */
      pre{
        background:#fbfbfc;                /* Light code background */
        border:1px solid #d9d9d9;          /* Subtle border */
        border-left:4px solid #0056b3;     /* Blue visual accent */
        padding:15px;}                     /* Internal spacing */
      code:not(.sourceCode):not(pre code){
        display:inline;                    /* Keep inline code in text */
        white-space:normal!important;      /* Allow text wrapping */
        word-wrap:break-word!important;    /* Prevent overflow */
        background:#eef3f8;                /* Highlight inline code */
        color:#002b7f;                     /* Match document palette */
        border:1px solid #d9d9d9;          /* Definition border */
        padding:2px 5px;}                  /* Compact padding */

      /* ---------- ABSTRACT ----------
         Academic summary box styling.
         Highlights introductory information
         while maintaining document consistency.
         ------------------------------------------ */
      .abstract{
        background:#f8f9fb;                    /* Neutral background */
        border-left:6px solid #002b7f;         /* Main theme accent */
        padding:18px 22px;                     /* Internal spacing */
        margin:25px 0;                         /* Separation from sections */
        color:#333;                            /* Softer text colour */
        line-height:1.7;                       /* Improve readability */
        text-align:justify;                    /* Academic alignment */
        box-shadow:0 2px 5px rgba(0,0,0,.05);} /* Subtle elevation */
      .abstract-title{
        color:#002b7f;                         /* Match heading colour */
        font-weight:700;                       /* Emphasize title */
        margin-bottom:10px;                    /* Space below title */
        text-transform:uppercase;              /* Editorial style */
        letter-spacing:.05em;}                 /* Improve readability */

      /* ---------- HEADINGS ----------
         Hierarchical styling for document sections.
         Uses colour, weight and separators to
         improve readability and navigation.
         ------------------------------------------ */
      h1{
        color:#002b7f;                    /* Main dark blue */
        padding-bottom:.3em;              /* Space before separator */
        margin-top:1.2em;                 /* Separation from previous content */
        font-weight:700;                  /* Strong emphasis */
        border-bottom:3px solid #002b7f;} /* Main section divider */
      h2{
        color:#004a99;                    /* Medium dark blue */
        font-weight:700;                  /* Strong subsection title */
        border-bottom:1px solid #0056b3;  /* Subtle separator */
        padding-bottom:5px;}              /* Spacing below title */
      h3{
        color:#0056b3;                    /* Secondary blue */
        font-weight:600;                  /* Medium emphasis */
        border-left:4px solid #0056b3;    /* Visual section marker */
        padding-left:10px;}               /* Space from marker */
      h4{color:#2f75b5;}                  /* Light professional blue */
      h5{color:#4f81bd;}                  /* Soft accent blue */
      h6{color:#7fa9d8;}                  /* Light blue for minor headings */

      /* ---------- TABLES ----------
         Professional formatting for statistical
         tables and generated outputs.
         ------------------------------------------ */
      table{
        width:100%;                 /* Use available width */
        border-collapse:collapse;}  /* Merge table borders */
      caption{
        caption-side:top;           /* Title above the table */
        color:#002b7f;              /* Main theme colour */
        font-weight:700;            /* Highlight table title */
        text-align:left;            /* Align with table content */
        padding-bottom:10px;}       /* Space below caption */
      th{
        background:#002b7f;         /* Header background */
        color:white;                /* High contrast text */
        padding:10px;               /* Cell spacing */
        border:1px solid #d9d9d9;}  /* Table definition */
      td{
        padding:8px 10px;           /* Comfortable reading space */
        border:1px solid #d9d9d9;}  /* Cell separation */
      tr:nth-child(even){
        background:#f7f9fb;}        /* Alternate row shading */
      tr:hover{background:#eef3f8;} /* Highlight active row */

      /* ---------- FIGURES ----------
         Image and caption styling for plots,
         diagrams and generated graphics.
         ------------------------------------------ */
      figure{
        margin:30px auto;                      /* Vertical spacing */
        text-align:center;                     /* Center figure content */
        max-width:100%;}                       /* Prevent overflow */
      figure img,
      img{
        max-width:95%;                         /* Keep margins around images */
        height:auto;                           /* Preserve proportions */
        border:1px solid #d9d9d9;              /* Subtle frame */
        padding:4px;                           /* Space around image */
        background:white;                      /* Clean background */
        box-shadow:0 1px 3px rgba(0,0,0,.08);} /* Light depth effect */
      figcaption,
      .figure-caption{
        color:#666;                            /* Secondary text colour */
        font-style:italic;                     /* Caption style */
        text-align:center;                     /* Center caption */
        font-size:.9rem;                       /* Smaller than body text */
        margin-top:8px;}                       /* Separation from image */

      /* ---------- MATH ----------
         Highlight mathematical expressions
         using the document accent colour.
         ------------------------------------------ */
      .math,
      mjx-container[jax="CHTML"],
      mjx-container[display="true"]{
        color:#7a1f2b;}       /* Bordeaux emphasis colour */
      .math{overflow-x:auto;} /* Support long equations */

      /* ---------- CALLOUTS ----------
         Harmonize Quarto callout boxes with
         the document colour palette.
         ------------------------------------------ */
      .callout{border-left-width:5px;}          /* Accent side indicator */
      .callout-note,
      .callout-tip{background:#f8f9fb;}         /* Neutral background */
      .callout-note{border-left-color:#0056b3;} /* Blue note accent */
      .callout-tip{border-left-color:#4f81bd;}  /* Lighter blue accent */
      .callout-warning{
        border-left-color:#b8860b;              /* Warning colour */
        background:#fffaf0;}                    /* Warm warning background */

      /* ---------- METADATA ----------
         Styling for title information,
         author data and document details.
         ------------------------------------------ */
      .title,
      .quarto-title-meta-heading{
        color:#002b7f;            /* Main document colour */
        font-weight:700;}         /* Emphasized metadata */
      .subtitle,
      .quarto-title-meta{
        color:#666;               /* Secondary information */
        font-style:italic;}       /* Distinguish metadata */

      /* ---------- OTHER ----------
         Additional document elements:
         separators, lists, references,
         quotations and form controls.
         ------------------------------------------ */
      hr{
        border:none;                   /* Remove default rule */
        border-top:2px solid #002b7f;  /* Main colour separator */
        margin:2em 0;}                 /* Vertical spacing */
      ul,
      ol{margin-bottom:1em;}           /* Space after lists */
      li{margin-bottom:.35em;}         /* Improve list readability */
      .footnotes{
        color:#666;                    /* Secondary text colour */
        font-size:.9rem;               /* Smaller footnote text */
        border-top:1px solid #d9d9d9;} /* Separation from content */
      blockquote{
        background:#f8f9fb;            /* Neutral background */
        border-left:5px solid #0056b3; /* Blue quotation marker */
        padding:12px 18px;             /* Internal spacing */
        color:#555;}                   /* Softer text colour */
      input,
      select{
        border:1px solid #d9d9d9;      /* Light form border */
        padding:6px;}                  /* Comfortable input spacing */

      /* ---------- MOBILE ----------
         Responsive adjustments for small screens.
         Improves readability and prevents
         overflow on mobile devices.
         ------------------------------------------ */
      @media(max-width:768px){
        body{
          font-size:15px;           /* Slightly smaller text */
          margin-top:20px;          /* Reduce top spacing */
          padding:0 12px;}          /* Add side margins */
        main{max-width:100%;}       /* Use full screen width */
        #toc,
        #contents{
          width:100%;               /* Expand TOC on mobile */
          margin:15px 0 25px;}      /* Adjust spacing */
        #TOC{position:relative;}    /* Preserve normal layout */
        p{text-align:left;}         /* Improve mobile reading */
        h1{font-size:1.7rem;}       /* Scale main headings */
        h2{font-size:1.35rem;}      /* Scale secondary headings */
        pre{
          padding:10px;             /* Reduce code spacing */
          font-size:.85rem;}        /* Smaller code text */
        table{
          display:block;            /* Enable horizontal scroll */
          overflow-x:auto;}
        img,
        figure img{max-width:100%;} /* Prevent image overflow */
      }

      </style>
---

# **Table of contents**

```{r, message=FALSE, warning=FALSE}
readLines("Quarto_Script.qmd") |> #read the Quarto document as a character vector
  (\(x) #anonymous function using the file lines without creating an object
   grep("^#{1,5}\\s*\\*\\*", #match headings from level 1 to 5 written in bold
      x, value = TRUE))() |> #return the matching lines instead of their indices
  (\(h) #anonymous function receiving the extracted heading lines
    cat(paste0(strrep("  ", #use two spaces for each indentation level
          nchar(sub("^(#{1,5}).*", #extract the sequence of leading '#' symbols
                    "\\1", h)) - 1), #level 1 has no indentation
          "- ", #prefix each heading with a bullet
        gsub("^#{1,5}\\s*\\*\\*|\\*\\*$", #remove markdown syntax
          "", h)), #keep only the heading text
        sep = "\n"))() #print one entry per line
```

# **Packages & Setup**

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
#to render the quarto document, press ctrl + shift + k.
#this code chunk initializes a clean r session to improve code reproducibility.

#This chunk initializes the R environment and sets global options to ensure full reproducibility of the analysis. It defines the CRAN mirror, configures download methods, and enforces IPv4 connectivity to avoid network issues. It also includes utilities to clean the workspace, detach loaded packages, remove existing objects, and reset the random seed, ensuring that the session starts from a completely clean state


#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#
#DOCUMENT PARAMETERS
#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#

#capture the document parameters (see the YAML `params` block) into session options before the workspace is cleared below. Session options survive environment resets and keep the dataset import configurable.
options(dq.data_path  = params$data_path,
        dq.data_type  = params$data_type,
        dq.data_delim = params$data_delim,
        dq.data_sheet = params$data_sheet)


#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#
#BEFORE STARTING
#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#

#set global r session options for (i) the cran repository used to install packages and (ii) the method used to download files from the internet.
options(repos = c(CRAN = "https://cran.stat.unipd.it/"), #specify the cran mirror used to download and install packages. using a fixed mirror improves reproducibility, download speed, and reliability.
  download.file.method = "libcurl") #use libcurl as the download method. it provides full https support, automatic redirects, and better stability than legacy methods.

#force libcurl to prefer ipv4 over ipv6; this can prevent connectivity issues on networks with unstable ipv6 routing or tls handshake failures.
Sys.setenv(CURL_IPRESOLVE = 1) #value 1 corresponds to curl_ipresolve_v4 (ipv4 only).

getwd() #display the current working directory.
#NOTE: do not hard-code an absolute working directory here. Paths are resolved relative to the
#R project root via here::here(), which keeps the workflow portable across machines and operating
#systems. To analyse a different dataset, set the `data_path` parameter in the YAML header (or pass
#it on render, e.g. quarto render Quarto_Script.qmd -P data_path:<path>) instead of using setwd().

# #restart the rstudio session.
# if (requireNamespace("rstudioapi", quietly = TRUE)) {rstudioapi::restartSession()}
# #this clears all loaded objects and packages while preserving the current script.


#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#
#CLEAN R ENVIRONMENT
#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#

#unload all attached packages except the core r system packages.
#search() returns all objects attached to the current r search path.
invisible(lapply(search(), function(x) {
  #check whether the current object is a package
  #(its name starts with "package:")
  #and ensure that it is not one of the essential base r packages.
  if (grepl("package:", x) &&
      !grepl("package:base", x) && #keep the base package.
      !grepl("package:stats", x) && #keep statistical functions.
      !grepl("package:utils", x) && #keep utility functions.
      !grepl("package:graphics", x) && #keep base graphics.
      !grepl("package:grDevices", x) && #keep graphics devices.
      !grepl("package:datasets", x) && #keep example datasets.
      !grepl("package:methods", x)) { #keep the methods package.
    #detach all non-system packages from the search path.
    #setting unload = true also unloads the package from memory.
    detach(x, character.only = TRUE, unload = TRUE)}}))

rm(list = ls(all.names = TRUE))
#remove all objects from the current workspace.
#ls(all.names = true) lists all objects, including hidden ones,
#and rm() deletes them.
#this ensures that the analysis starts from a clean environment.
#(the dataset path/separator were already stored in options() above, so they
#survive this reset and the later garbage-cleaning chunks.)

rm(".Random.seed", envir = .GlobalEnv) #remove any previously defined random seed.

# set.seed(42) #set a reproducible random seed.
# #print the current random seed if it exists.
# if (exists(".Random.seed")) {
#   cat("\ncurrent random seed:\n")
#   print(.Random.seed)
# } else {
#   cat("\nno random seed has been generated in the current session.\n")}

gc() #trigger garbage collection to release unused memory.

graphics.off() #close all open graphics devices to avoid conflicts with new plots.

closeAllConnections() #close all open connections to prevent resource conflicts.

#clear the r console (equivalent to pressing ctrl + l in rstudio).
cat("\014") #"\014" is the form-feed character interpreted by rstudio as "clear console".
```

```{r setup-time, include=FALSE, echo=FALSE}
#This chunk creates an isolated environment to store the execution start time of the script. It is used to measure total runtime and support reproducibility tracking by recording when the analysis begins.

#create an isolated environment, that is, an empty environment, useful for storing temporary objects without cluttering the global environment.
.time_env <- new.env(parent = emptyenv())

#store the execution start timestamp. Sys.time() returns the current date and time.
.time_env$start <- Sys.time()
```

```{r, message=FALSE, warning=FALSE}
#| code-fold: true 
#| code-summary: "Parallel processing"

#This chunk configures parallel computing capabilities using multiple CPU cores. It installs and loads the required packages and defines a multisession plan that uses all available cores minus one. This improves computational efficiency for operations that support parallel execution.

#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#
#PARALLEL PROCESSING
#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#---#

if (!requireNamespace("parallel", quietly = TRUE)){install.packages("parallel")}
library(parallel) #functions for native parallel computing in R.
# packageDescription("parallel")

if (!requireNamespace("furrr", quietly = TRUE)){install.packages("furrr")}
library(furrr) #extends purrr to execute functions in parallel using future.
# packageDescription("furrr")

plan(multisession, workers = parallel::detectCores() - 1)
#set the parallel execution plan for future/furrr.
#'multisession' creates multiple R sessions running in parallel
#(on Windows, each session runs as a separate process).
#'workers = parallel::detectCores() - 1' uses all available CPU cores except one,
#preventing the CPU from being fully saturated and leaving resources available for the operating system.
```

```{r, message=FALSE, warning=FALSE}
#| code-fold: true 
#| code-summary: "Packages"

#This chunk installs and loads all necessary packages for data analysis, visualization, statistical computation, and reporting. It includes tools for data manipulation (`tidyverse`), descriptive statistics (`psych`), association measures (`rcompanion`), and HTML reporting (`knitr`, `kableExtra`), ensuring that all required dependencies are available for the workflow.

# Import data
if (!requireNamespace("readr", quietly = TRUE)) {install.packages("readr")}
library(readr) #import CSV files.
# packageDescription("readr")

if (!requireNamespace("readxl", quietly = TRUE)) {install.packages("readxl")}
library(readxl) #import Excel files.
# packageDescription("readxl")

# Data manipulation and management
if (!requireNamespace("tidyverse", quietly = TRUE)) {install.packages("tidyverse")}
#tidyverse is a meta-package for data manipulation, analysis, and visualization, including:
#- ggplot2 for data visualization using graphics.
#- dplyr for data manipulation, with functions such as filter(), select(), summarise(), and mutate().
#- tidyr for reshaping data between wide and long formats.
#- readr for importing CSV and text files.
#- purrr for functional programming and iteration using functions such as map().
#- tibble for an improved and modern implementation of data frames.
#- stringr for string manipulation.
#- forcats for handling and reordering categorical variables (factors).
#- lubridate for working with dates and times.
library(tidyverse) #data manipulation, analysis, and visualization.
# packageDescription("tidyverse")

# Data cleaning and descriptive statistics
if (!requireNamespace("janitor", quietly = TRUE)){install.packages("janitor")}
library(janitor) #clean and standardize column names.
# packageDescription("janitor")

if (!requireNamespace("psych", quietly = TRUE)){install.packages("psych")}
library(psych) #compute descriptive statistics using describe().
# packageDescription("psych")

if (!requireNamespace("skimr", quietly = TRUE)){install.packages("skimr")}
library(skimr) #compact summary statistics via skim().
# packageDescription("skimr")

if (!requireNamespace("rcompanion", quietly = TRUE)){install.packages("rcompanion")}
library(rcompanion) #compute Cramér's V.
# packageDescription("rcompanion")

# HTML output and reporting
if (!requireNamespace("knitr", quietly = TRUE)){install.packages("knitr")}
library(knitr) #dynamic reporting with R Markdown and Quarto.
# packageDescription("knitr")

if (!requireNamespace("kableExtra", quietly = TRUE)){install.packages("kableExtra")}
library(kableExtra) #create and customize HTML tables.
# packageDescription("kableExtra")
```

<details> 
<summary>Show complete reproducibility information</summary>

```{r, message=FALSE, warning=FALSE}
#| code-fold: true 
#| code-summary: "Code for reproducibility information"

#This chunk prints a detailed report of the computing environment, including operating system, hardware specifications, memory usage, R version, installed tools, and session configuration. It also records random number generator settings and file system information to ensure full reproducibility of the analysis across different machines and sessions.

cat(" =========================================\n",
    " REPRODUCIBILITY INFORMATION\n",
    "=========================================\n\n",
    
    #this section collects essential information about the hardware environment and operating system.
    "=== SYSTEM AND HARDWARE ===\n",
    
    "Operating system:\n",
    #operating system name (windows, linux, macos)
    "    OS name:                          ", Sys.info()["sysname"], "\n",
    #specific os version
    "    OS version:                       ", Sys.info()["release"], "\n",
    #machine architecture
    "    Architecture:                     ", Sys.info()["machine"], "\n",
    #number of logical cpu cores available for parallel computation
    "    CPU cores (logical):              ", parallel::detectCores(), "\n",
    #information about r compilation platform
    "    R compilation platform:           ", R.version$platform, "\n\n",
    
    #blas and lapack are fundamental mathematical libraries for linear algebra and numerical computing in r;
    #here the version of external linked libraries is printed
    "BLAS/LAPACK:\n",
    paste0("    ",names(extSoftVersion()), ": ", extSoftVersion(),collapse = "\n "), "\n\n",

    #this section describes available memory and current r session usage
    "Memory:\n",
    
    #estimated total system ram (in gb)
    "    Total RAM (GB): ", 
    if (.Platform$OS.type == "windows") {
     #on windows use wmic command
      as.numeric(gsub("[^0-9]", "", 
                      system("wmic computersystem get TotalPhysicalMemory",intern = TRUE)[2])) / (1024^3)
    } else if (Sys.info()["sysname"] == "Linux") {
     #on linux read from /proc/meminfo
      as.numeric(system("grep MemTotal /proc/meminfo | awk '{print $2}'",intern = TRUE)) / (1024^2)
    } else if (Sys.info()["sysname"] == "Darwin") {
     #on macos use sysctl
      as.numeric(system("sysctl -n hw.memsize",intern = TRUE)) / (1024^3)
    } else {NA},"\n",

    #r garbage collector automatically manages memory by freeing unused objects; these values show current usage and internal memory limits during the session
    "    Garbage collector (R memory management):\n",
    paste(sprintf("        %-7s | used: %8s | trigger: %8s | max used: %8s",
                  rownames(gc()),
                  format(gc()[, "used"], big.mark = ","),
                  format(gc()[, "gc trigger"], big.mark = ","),
                  format(gc()[, "max used"], big.mark = ",")),
          collapse = "\n "),"\n\n",
    
    #information about r version and related tools
    "=== R ENVIRONMENT ===\n",
    
    "R:\n",
    #full r version string
    "    Version:                         ", R.version.string, "\n",
    #r installation directory
    "    R installation directory:        ", R.home(), "\n\n",

    "RStudio:\n",
    #check whether rstudio is available and display its version
    "    Version:                         ",
    if (requireNamespace("rstudioapi", quietly = TRUE) &&
        rstudioapi::isAvailable()) {
      as.character(rstudioapi::getVersion())
    } else {
      "Not available, running outside RStudio"
    },
    "\n\n",

    "Quarto:\n",
    #quarto version, if installed on the system
    "    Version:                         ",
    tryCatch(system("quarto --version", intern = TRUE),
             error = function(e) "Not available"),
    "\n\n",

    #information about locale, numeric format, and date/time settings
    "=== LOCALIZATION AND TIME ===\n",
    
    "Localization:\n",
    "    Collate:   ", Sys.getlocale("LC_COLLATE"), "\n",
    "    Ctype:     ", Sys.getlocale("LC_CTYPE"), "\n",
    "    Monetary:  ", Sys.getlocale("LC_MONETARY"), "\n",
    "    Numeric:   ", Sys.getlocale("LC_NUMERIC"), "\n",
    "    Time:      ", Sys.getlocale("LC_TIME"), "\n\n",

    "Date and time:\n",
    #session timestamp
    "    Timestamp: ", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), "\n",
    #system timezone
    "    Timezone:  ", Sys.timezone(), "\n\n",

    #essential for reproducibility of statistical analyses, since changing the seed or rng method changes results
    "=== RANDOM NUMBER GENERATOR ===\n",
    
    "    RNG kind:      ", paste(RNGkind(), collapse = ", "), "\n",
    #check whether a seed exists in the current session
    "    Seed present:  ", if (exists(".Random.seed")) "YES" else "NO", "\n",
    #print current seed if available
    "    Current seed:  ", if (exists(".Random.seed")) {paste(.Random.seed, collapse = " ")
      } else {"No seed present"}, "\n\n",

    #useful for project traceability (git + quarto files)
    "=== VERSIONING AND CODE ===\n",
    
    "Git:\n",
    #git commit hash (if available)
    "    Commit: ",tryCatch(system("git rev-parse HEAD", intern = TRUE),
                            error = function(e) "Not initialized or repository not found"),"\n",
    #current git branch
    "    Branch: ",tryCatch(system("git branch --show-current", intern = TRUE),
                            error = function(e) "Not initialized or repository not found"),"\n\n",
    
    "Quarto code files (.qmd):\n",
    #number of quarto files in directory and subdirectories
    "    Number of files: ", length(list.files(pattern = "\\.qmd$", recursive = TRUE)), "\n",
    #md5 hashes of quarto files for integrity/version control
    #md5 (message digest algorithm 5) is a hash function that generates a unique fingerprint of a file.
    #even a small change produces a completely different hash.
    #it is used to verify file integrity and detect modifications.
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.qmd$", recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.qmd$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "RMarkdown files (.Rmd):\n",
    #number of rmarkdown files
    "    Number of files: ", length(list.files(pattern = "\\.Rmd$", recursive = TRUE)), "\n",
    #md5 hashes of rmarkdown files
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.Rmd$", recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.Rmd$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "R scripts (.R):\n",
    #number of r files
    "    Number of files: ", length(list.files(pattern = "\\.R$", recursive = TRUE)), "\n",
    #md5 hashes of r files
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.R$", recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.R$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "Jupyter notebooks (.ipynb):\n",
    #number of jupyter files
    "    Number of files: ", length(list.files(pattern = "\\.(ipynb)$", recursive = TRUE)), "\n",
    #md5 hashes
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.(ipynb)$", recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.(ipynb)$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "Python scripts (.py):\n",
    #number of python files
    "    Number of files: ", length(list.files(pattern = "\\.(py)$", recursive = TRUE)), "\n",
    #md5 hashes
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.(py)$", recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.(py)$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "MATLAB files (.m):\n",
    #number of matlab files
    "    Number of files: ", length(list.files(pattern = "\\.m$", recursive = TRUE)), "\n",
    #md5 hashes
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.m$", recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.m$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "SPSS files (.sps):\n",
    #number of spss files
    "    Number of files: ", length(list.files(pattern = "\\.sps$", recursive = TRUE)), "\n",
    #md5 hashes
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.sps$", recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.sps$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "Stata files (.do):\n",
    #number of stata files
    "    Number of files: ", length(list.files(pattern = "\\.do$", recursive = TRUE)), "\n",
    #md5 hashes
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.do$", recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.do$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "Excel files (.xls, .xlsx, .xlsm):\n",
    #number of excel files
    "    Number of files: ", length(list.files(pattern = "\\.(xls|xlsx|xlsm)$", recursive = TRUE)), "\n",
    #md5 hashes
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.(xls|xlsx|xlsm)$", 
                                                               recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.(xls|xlsx|xlsm)$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "SAS files (.sas):\n",
    #number of sas files
    "    Number of files: ", length(list.files(pattern = "\\.(sas)$", recursive = TRUE)), "\n",
    #md5 hashes
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.(sas)$", 
                                                               recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.(sas)$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "JASP files (.jasp):\n",
    #number of jasp files
    "    Number of files: ", length(list.files(pattern = "\\.jasp$", recursive = TRUE)), "\n",
    #md5 hashes
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.jasp$", recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.jasp$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "Jamovi files (.omv):\n",
    #number of jamovi files
    "    Number of files: ", length(list.files(pattern = "\\.omv$", recursive = TRUE)), "\n",
    #md5 hashes
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.omv$", recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.omv$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "Power BI files (.pbix):\n",
    #number of power bi files
    "    Number of files: ", length(list.files(pattern = "\\.pbix$", recursive = TRUE)), "\n",
    #md5 hashes
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.pbix$", recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.pbix$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    #input dataset tracking
    "=== INPUT DATA ===\n",
    
    "Tabular data files (.csv, .tsv, .tab, .psv, .dat, .fwf, .txt, .xls, .xlsx, .xlsm, .ods, .numbers):\n",
    "    Number of files: ", length(list.files(pattern = "\\.(csv|tsv|tab|psv|dat|fwf|txt|xls|xlsx|xlsm|ods|numbers)$", recursive = TRUE)), "\n",
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.(csv|tsv|tab|psv|dat|fwf|txt|xls|xlsx|xlsm|ods|numbers)$", 
                                                               recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.(csv|tsv|tab|psv|dat|fwf|txt|xls|xlsx|xlsm|ods|numbers)$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "Structured data files (.json, .ndjson, .xml, .yaml, .yml):\n",
    "    Number of files: ", length(list.files(pattern = "\\.(json|ndjson|xml|yaml|yml)$", recursive = TRUE)), "\n",
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.(json|ndjson|xml|yaml|yml)$", 
                                                               recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.(json|ndjson|xml|yaml|yml)$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "R data files (.rds, .rda, .RData):\n",
    "    Number of files: ", length(list.files(pattern = "\\.(rds|rda|RData)$", recursive = TRUE)), "\n",
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.(rds|rda|RData)$", 
                                                               recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.(rds|rda|RData)$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "Database files (.sqlite, .sqlite3, .db, .duckdb, .mdb, .accdb, .sql):\n",
    "    Number of files: ", length(list.files(pattern = "\\.(sqlite|sqlite3|db|duckdb|mdb|accdb|sql)$", recursive = TRUE)), "\n",
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.(sqlite|sqlite3|db|duckdb|mdb|accdb|sql)$", 
                                                               recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.(sqlite|sqlite3|db|duckdb|mdb|accdb|sql)$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "Statistical/legacy data files (.sav, .zsav, .por, .dta, .sas7bdat, .xpt, .sd2, .sd7, .mtp, .jmp, .wf1, .gdt):\n",
    "    Number of files: ", length(list.files(pattern = "\\.(sav|zsav|por|dta|sas7bdat|xpt|sd2|sd7|mtp|jmp|wf1|gdt)$", recursive = TRUE)), "\n",
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.(sav|zsav|por|dta|sas7bdat|xpt|sd2|sd7|mtp|jmp|wf1|gdt)$", 
                                                               recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.(sav|zsav|por|dta|sas7bdat|xpt|sd2|sd7|mtp|jmp|wf1|gdt)$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "Big data/scientific files (...):\n",
    "    Number of files: ", length(list.files(pattern = "\\.(feather|arrow|ipc|parquet|orc|avro|delta|iceberg|h5|hdf5|nc|fits|nii|mat|npy|npz|pkl|pickle|joblib)$", recursive = TRUE)), "\n",
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.(feather|arrow|ipc|parquet|orc|avro|delta|iceberg|h5|hdf5|nc|fits|nii|mat|npy|npz|pkl|pickle|joblib)$", 
                                                               recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.(feather|arrow|ipc|parquet|orc|avro|delta|iceberg|h5|hdf5|nc|fits|nii|mat|npy|npz|pkl|pickle|joblib)$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "Geospatial data files (.shp, .geojson, .gpkg, .kml):\n",
    "    Number of files: ", length(list.files(pattern = "\\.(shp|geojson|gpkg|kml)$", recursive = TRUE)), "\n",
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.(shp|geojson|gpkg|kml)$", 
                                                               recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.(shp|geojson|gpkg|kml)$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    "Archive/dashboard files (.zip, .gz, .bz2, .xz, .7z, .pbix, .hyper):\n",
    "    Number of files: ", length(list.files(pattern = "\\.(zip|gz|bz2|xz|7z|pbix|hyper)$", recursive = TRUE)), "\n",
    tryCatch(paste(sprintf("    %s | %s",
                           names(tools::md5sum(sort(list.files(pattern = "\\.(zip|gz|bz2|xz|7z|pbix|hyper)$", 
                                                               recursive = TRUE)))),
                           tools::md5sum(sort(list.files(pattern = "\\.(zip|gz|bz2|xz|7z|pbix|hyper)$", recursive = TRUE)))),
                   collapse = "\n "), error = function(e) "    Not available"), "\n\n",
    
    #full r session description
    "=== R SESSION ===\n",
    
    "Loaded packages (with dependencies):\n",
    paste(sapply(sort(grep("^package:", search(), value = TRUE)),
                 function(pkg) paste0("    ", sub("^package:", "", pkg),
                                      " (", packageVersion(sub("^package:", "", pkg)), ")\n",
                                      "        Depends: ",gsub("\n", " ",
                                                              tryCatch(packageDescription(
                                                                sub("^package:", "", pkg))$Depends,
                                                                error = function(e) "none")),"\n",
                                      "        Imports: ",gsub("\n", " ",
                                                              tryCatch(packageDescription(
                                                                sub("^package:", "", pkg))$Imports,
                                                                error = function(e) "none")))),
          collapse = "\n"),"\n\n",

    #library paths where r installs packages
    "Library paths:\n",
    paste0("    ", .libPaths(), collapse = "\n "),"\n\n",

    #search path elements that are not packages
    "Search path (non-package):\n",
    if(length(setdiff(search(),
                      grep("^package:", search(), value = TRUE))) > 0) {
      paste0("    ",paste(sort(setdiff(search(),grep("^package:", search(), value = TRUE))),
                          collapse = "\n     "))
    } else {"    None"},"\n\n",
    
    #objects currently in memory
    "Objects in memory:\n",
    "    Count:             ", length(ls()), "\n",
    
    #total memory used by objects
    if(length(ls()) > 0) {
      paste0("    Total memory (KB): ",
             round(sum(sapply(ls(), function(x)
               tryCatch(as.numeric(object.size(get(x))),
                        error = function(e) 0))) / 1024, 2))
    } else {"    0 KB used (no objects in memory)"},"\n",
    
    #list objects with their class
    if(length(ls()) > 0) {
      paste0("    List:\n",
             paste0("        ",ls(), " (",sapply(ls(), function(x) class(get(x))[1]),")",
                    collapse = "\n"))
    } else {"    No objects"},"\n\n",
    
    #working directory and contents
    "=== FILE SYSTEM ===\n",
    "Working directory:\n",
    "    Path:        ", getwd(), "\n",
    #total number of files in working directory and subdirectories
    "    Number of files: ", length(list.files(path = getwd(), recursive = TRUE)), "\n",
    "    Directory contents (including subdirectories):\n",
    #detailed file listing with type, extension, and path
    if (length(list.files(path = getwd(), recursive = TRUE)) > 0) {
      paste0("        ", 
             sprintf("%s | type: %s | size: %.3f KB | path: %s", 
                     list.files(path = getwd(), recursive = TRUE),
                     ifelse(tools::file_ext(list.files(path = getwd(), recursive = TRUE)) == "", 
                            "directory", tools::file_ext(list.files(path = getwd(), recursive = TRUE))),
                     file.info(list.files(path = getwd(), recursive = TRUE))$size / 1024,
                     dirname(list.files(path = getwd(), recursive = TRUE))),
             collapse = "\n ")
      } else {"       Empty directory"},"\n\n",

    #final separator for readability
    "=========================================\n")
```

</details>

# **The dataset**

This template uses a real-world dataset from the OpenPsychometrics project, which contains responses used to compute the Big Five personality traits. The data are freely available and are suitable for demonstrating data cleaning, descriptive statistics, and multivariate analysis workflows.

The dataset can be downloaded from the following source (Open Psychometrics Project, n.d.):

[http://openpsychometrics.org/_rawdata/BIG5.zip](http://openpsychometrics.org/_rawdata/BIG5.zip)

```{r}
#link where is possible downloaded data used for example
url <- "http://openpsychometrics.org/_rawdata/BIG5.zip"

#name of the local file where the dataset will be saved
destfile <- "BIG5.zip"

#download the ZIP file from the specified URL
download.file(url, destfile = destfile, mode = "wb")

#extract the contents of the ZIP archive into a folder
unzip(destfile, exdir = "BIG5_data")
```

## **Data Dictionary**

The data were collected around 2012 through an interactive online personality assessment. Participants were informed at the beginning of the test that their responses would be recorded and used for research purposes. At the end of the questionnaire, participants were asked to confirm their consent for the use of their data in research. Only participants who provided consent were included in the dataset.

Participants completed a personality questionnaire based on the Big Five personality traits. Each item was rated on a 5-point Likert scale:

| Value | Meaning          |
| ----- | ---------------- |
| 1     | Disagree         |
| 3     | Neutral          |
| 5     | Agree            |
| 0     | Missing response |

Items were presented on a single page in the following order: `E1, N2, A1, C1, O1, E2, ...`

### Extraversion (E)

| Code | Item                                            |
| ---- | ----------------------------------------------- |
| E1   | I am the life of the party.                     |
| E2   | I don't talk a lot.                             |
| E3   | I feel comfortable around people.               |
| E4   | I keep in the background.                       |
| E5   | I start conversations.                          |
| E6   | I have little to say.                           |
| E7   | I talk to a lot of different people at parties. |
| E8   | I don't like to draw attention to myself.       |
| E9   | I don't mind being the center of attention.     |
| E10  | I am quiet around strangers.                    |

### Neuroticism (N)

| Code | Item                           |
| ---- | ------------------------------ |
| N1   | I get stressed out easily.     |
| N2   | I am relaxed most of the time. |
| N3   | I worry about things.          |
| N4   | I seldom feel blue.            |
| N5   | I am easily disturbed.         |
| N6   | I get upset easily.            |
| N7   | I change my mood a lot.        |
| N8   | I have frequent mood swings.   |
| N9   | I get irritated easily.        |
| N10  | I often feel blue.             |

### Agreeableness (A)

| Code | Item                                            |
| ---- | ----------------------------------------------- |
| A1   | I feel little concern for others.               |
| A2   | I am interested in people.                      |
| A3   | I insult people.                                |
| A4   | I sympathize with others' feelings.             |
| A5   | I am not interested in other people's problems. |
| A6   | I have a soft heart.                            |
| A7   | I am not really interested in others.           |
| A8   | I take time out for others.                     |
| A9   | I feel others' emotions.                        |
| A10  | I make people feel at ease.                     |

### Conscientiousness (C)

| Code | Item                                                     |
| ---- | -------------------------------------------------------- |
| C1   | I am always prepared.                                    |
| C2   | I leave my belongings around.                            |
| C3   | I pay attention to details.                              |
| C4   | I make a mess of things.                                 |
| C5   | I get chores done right away.                            |
| C6   | I often forget to put things back in their proper place. |
| C7   | I like order.                                            |
| C8   | I shirk my duties.                                       |
| C9   | I follow a schedule.                                     |
| C10  | I am exacting in my work.                                |

### Openness to Experience (O)

| Code | Item                                            |
| ---- | ----------------------------------------------- |
| O1   | I have a rich vocabulary.                       |
| O2   | I have difficulty understanding abstract ideas. |
| O3   | I have a vivid imagination.                     |
| O4   | I am not interested in abstract ideas.          |
| O5   | I have excellent ideas.                         |
| O6   | I do not have a good imagination.               |
| O7   | I am quick to understand things.                |
| O8   | I use difficult words.                          |
| O9   | I spend time reflecting on things.              |
| O10  | I am full of ideas.                             |

### Demographic Variables

The following demographic information was collected on a separate page:

| Variable | Description                                                                       |
| -------- | --------------------------------------------------------------------------------- |
| `race`   | Self-reported ethnic background selected from a predefined list of 13 categories. |
| `age`    | Age entered as free text. Participants younger than 13 years were excluded.       |
| `engnat` | Native English speaker (`1 = Yes`, `2 = No`, `0 = Missing`).                      |
| `gender` | Gender (`1 = Male`, `2 = Female`, `3 = Other`, `0 = Missing`).                    |
| `hand`   | Writing hand preference (`1 = Right`, `2 = Left`, `3 = Both`, `0 = Missing`).     |

Participants were also asked to confirm that their responses were accurate and could be used for research purposes. Records without confirmation were discarded.

### Technical Variables

Additional variables were derived from technical metadata collected during participation:

| Variable  | Description                                                           |
| --------- | --------------------------------------------------------------------- |
| `country` | Participant's estimated location represented by the ISO country code. |
| `source`  | Traffic source inferred from the HTTP Referrer field.                 |

The `source` variable was encoded as follows:

| Code | Source                                       |
| ---- | -------------------------------------------- |
| 1    | Another page within the test website         |
| 2    | Google                                       |
| 3    | Facebook                                     |
| 4    | Educational domain (`.edu`, `.edu.au`, etc.) |
| 6    | Other source or missing HTTP Referrer        |

## **First exploration of the dataset**

### **Data import and dataset dimensions**

```{r}
#retrieve the dataset import parameters from the session options
data_path  <- getOption("dq.data_path")
data_type  <- tolower(getOption("dq.data_type"))
data_delim <- getOption("dq.data_delim")
data_sheet <- getOption("dq.data_sheet")

#resolve relative paths from the project root
if (!grepl("^(/|[A-Za-z]:)", data_path)) {data_path <- here::here(data_path)}

#remove all objects from the environment except those explicitly required
rm(list = setdiff(ls(),
                  c("data_0",
                    "data_path",
                    "data_type",
                    "data_delim",
                    "data_sheet")))

#import the dataset according to the selected file type
if (data_type == "csv") {
  #read delimited text files using the specified column separator
  data_0 <- readr::read_delim(data_path,
                              delim = data_delim, #column separator (e.g. "," ";" or "\t")
                              escape_double = FALSE,
                              trim_ws = TRUE)
  } else if (data_type == "excel") {
    #read the selected worksheet from an Excel workbook
    data_0 <- readxl::read_excel(data_path,
                                 sheet = data_sheet) #worksheet name or index
    } else {stop("Unsupported data_type. Use 'csv' or 'excel'.")}

#print dataset dimensions
cat("The dataset has", nrow(data_0), "rows and", ncol(data_0), "columns\n")
```

#### **Renaming variables**

```{r}
original_names <- names(data_0) #store the original variable names for later comparison

#convert variable names to snake_case style.
#snake_case is a naming convention where all letters are lowercase and words are separated by underscores (_)
data_1 <- data_0 %>% janitor::clean_names() 

new_names <- names(data_1) #store the new snake_case variable names for comparison

#create a mapping table
mapping <- tibble(`Original name` = original_names, `Snake_case name` = new_names)

#display mapping table in HTML format
mapping %>%
  select(`Original name`, `Snake_case name`) %>%
  #create an HTML table
  kbl("html", caption = "Mapping between original and renamed variables",
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")

#remove all objects except the dataset
rm(list = setdiff(ls(), c("data_1")))
```

### **Dataset overview**

```{r}
set.seed(42) #set a seed for reproducibility

data_1 %>% 
 #randomly select 6 observations; results depend on the seed
  slice_sample(n = 6) %>% 
  #display the dataset as an HTML table
  kbl("html", caption = "6 random observations from the dataset",
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")
```

### **Missing or invalid values**

#### **Check by variable**

```{r}
data_1 %>%
  #convert everything to character; this is to standardize the handling of values in subsequent operations
  mutate(across(everything(), as.character)) %>%
  #reshape the dataset from wide to long format, with each row representing an observed value of a variable
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  #group by variable to compute different types of missing or invalid values
  group_by(Variable) %>%
  summarise(NA_count = sum(is.na(Value)), #true NA values
            empty_count = sum(Value == "", na.rm = TRUE), #empty strings
            space_count = sum(Value == " ", na.rm = TRUE), #strings containing only a space
            #toupper converts any combination of uppercase and lowercase letters to uppercase
            NONE_count = sum(toupper(Value) == "NONE", na.rm = TRUE), #textual "NONE" values 
            NULL_count = sum(toupper(Value) == "NULL", na.rm = TRUE), #textual "NULL" values
            NAN_count = sum(toupper(Value) == "NAN", na.rm = TRUE), #textual "NAN" values
            total_obs = n(), #total observations per variable
            .groups = "drop") %>% #remove grouping
  #compute total invalid values and their percentage over total observations
  mutate(total_invalid = NA_count + empty_count + space_count + NONE_count + NULL_count + NAN_count,
         #compute percentage of invalid values over total and round to 3 decimals
         perc_invalid = round((total_invalid / total_obs) * 100, 3)) %>%
  arrange(desc(total_invalid)) %>% #sorted in descending order, from most to least problematic
  #convert to HTML table
  kbl("html", caption = "Missing or invalid values by variable", 
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")
```

#### **Check by observation**

```{r}
data_1 %>%
  #convert all variables to character to avoid issues with non-standard missing value encodings
  mutate(across(everything(), as.character)) %>%
  #compute, for each observation (row), the number of invalid values
  mutate(n_invalid = rowSums(is.na(.) | #true NA values OR
                               . == "" | #empty strings OR
                               . == " " | #single spaces OR
                               toupper(.) == "NONE" | #missing values encoded as "NONE" OR
                               toupper(.) == "NULL" | #missing values encoded as "NULL" OR
                               toupper(.) == "NAN")) %>% #missing values encoded as "NAN"
  #build the distribution of the number of invalid values per observation
  count(n_invalid) %>% #how many rows have 0, 1, 2, ... invalid values
  arrange(n_invalid) %>% #sort in ascending order, starting from rows without missing values
  #rename columns to make the table more readable
  rename(`Number of invalid per observation` = n_invalid,
         `Number of observations` = n) %>%
  #convert to HTML table
  kbl("html", caption = "Distribution of missing or invalid values per observation", 
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")
```

```{r}
data_1 %>%
  mutate(across(everything(), as.character)) %>% #convert all variables to character to avoid type issues
  mutate(row_id = row_number()) %>% #create a sequential identifier for each row to track observations
  #reshape the dataset from wide to long format, where each cell becomes a row
  tidyr::pivot_longer(cols = -c(row_id),
                      names_to = "item",
                      values_to = "value") %>%
  #define as "invalid" values that are missing, empty, or encoded as NONE/NULL/NAN
  mutate(invalid = is.na(value) |
           value == "" |
           value == " " |
           toupper(value) %in% c("NONE", "NULL", "NAN")) %>%

  filter(invalid) %>% #keep only cells with invalid values
  group_by(row_id) %>% #group by observation using the original row identifier
  summarise(`Number of invalid values` = n(), #count how many invalid values each row has
            `Incomplete variables` = paste(item, collapse = ", "), #list variables with issues
            .groups = "drop") %>% #remove grouping after summarisation
  arrange(desc(`Number of invalid values`)) %>% #sort observations from most to least problematic
  #create formatted HTML table
  kbl("html", caption = "Participants with missing or invalid values and corresponding variables", 
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")
```

### **Non-informative variables**

```{r}
data_1 %>%
  #convert everything to character to work with unique values
  mutate(across(everything(), as.character)) %>%
  #compute, for each variable, the number of unique values if the values are less than or equal to 100, they are also printed
  summarise(across(everything(), 
                   list(Num_unique_values = ~ n_distinct(.),
                        Values = ~ {u <- unique(na.omit(.)) #remove non-informative NA values
                        #if the number of levels is below 100, print them; otherwise indicate it
                        if (length(u) <= 100) {paste(u, collapse = ", ")
                          } else {"unique values > 100"}}))) %>%
  #reshape the dataset from wide to long format to obtain one row per variable with its number of levels and values
  pivot_longer(everything(),
               names_to = c("Variable", ".value"),
               names_pattern = "^(.*)_(Num_unique_values|Values)$") %>%
  arrange(Num_unique_values) %>% #sort in ascending order by number of unique values
  #create formatted HTML table
  kbl("html", caption = "Number of unique values per variable", 
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")
```

#### **Removal of degenerate (constant) variables**

```{r}
#print variables that are entirely NA and variables with only one unique value
cat("All-NA variables:\n",
    paste(names(select(data_1, where(~ all(is.na(.))))), collapse = "\n "),
    "\n\nConstant variables:\n",
    paste(names(select(data_1, where(~ n_distinct(., na.rm = FALSE) == 1))), collapse = "\n "),
    "\n\n")

data_2 <- data_1 %>%
  select(where(~ !all(is.na(.)))) %>% #keep only columns that are not entirely NA
  select(where(~ n_distinct(., na.rm = FALSE) > 1)) #keep only non-constant columns (with more than one unique value)

#print dataset dimensions
cat("The dataset has", nrow(data_2), "rows and", ncol(data_2), "columns\n")

rm(list = setdiff(ls(), c("data_2")))
```

### **Type of variables**

```{r}
#compact structure of the object:
#number of rows and columns of the dataset
#column names, type of each column, and first observations of data_2
glimpse(data_2)
```

```{r}
data_2 %>%
  select(everything()) %>% #select all variables in the dataset of interest
  #convert to character to avoid implicit coercion and to manually identify non-numeric values
  mutate(across(everything(), as.character)) %>%
  #compute data quality metrics for each variable
  summarise(across(everything(), 
                   list(
                     #count of values that cannot be converted to numeric, 
                     #where coercion produces NA indicating non-numeric values
                     Num_non_numeric_values = ~ {
                       v <- . #include original NAs, which are excluded in the following sum
                       sum(!is.na(v) & is.na(suppressWarnings(as.numeric(v))))},
                     #percentage of non-numeric values over total observations
                     Perc_non_numeric_values = ~ {
                       v <- . #include original NAs, which are excluded in the numerator of the percentage
                       sum(!is.na(v) & is.na(suppressWarnings(as.numeric(v)))) / length(v) * 100},
                     #extract observed non-numeric values to diagnose import or encoding issues
                     Non_numeric_values = ~ {
                       v <- na.omit(.) #exclude structural missing values
                       #select only values that fail numeric conversion
                       u <- unique(v[is.na(suppressWarnings(as.numeric(v)))])
                       #if fewer than 100 values, display them explicitly; otherwise avoid long output
                       if (length(u) <= 100) {paste(u, collapse = ", ")
                       } else {"non-numeric values > 100"}}))) %>%
  #reshape from wide to long format to obtain one row per variable with all associated metrics
  pivot_longer(everything(),
               names_to = c("Variable", ".value"),
               names_pattern = "^(.*)_(Num_non_numeric_values|Perc_non_numeric_values|Non_numeric_values)$") %>%
  #round percentage to three decimals for readability
  mutate(Perc_non_numeric_values = round(Perc_non_numeric_values, 3)) %>%
  #sort variables from most to least problematic
  arrange(desc(Num_non_numeric_values)) %>%
  #output as HTML table
  kbl("html", caption = "Non-numeric values per variable", 
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")
```

```{r}
data_2 %>%
  select(everything()) %>% #select all variables in the dataset of interest
  #convert all variables to character to avoid implicit coercion and to manually handle numeric conversion in subsequent steps
  mutate(across(everything(), as.character)) %>%
  #compute descriptive metrics on integer numeric values for each variable
  summarise(across(
    everything(),
    list(
      #count of integer numeric values present in the variable
      Num_integer_values = ~ {
        #attempt numeric conversion while suppressing warnings
        #generated by non-convertible values
        v <- suppressWarnings(as.numeric(.))
        #count only non-missing values after conversion that have zero remainder in division by 1, i.e., no decimal component
        sum(!is.na(v) & v %% 1 == 0)},
      #compute percentage of integer numeric values over total observations in the variable
      Perc_integer_values = ~ {
        v <- suppressWarnings(as.numeric(.)) #convert variable to numeric
        #compute percentage share of integer values relative to total variable length
        sum(!is.na(v) & v %% 1 == 0) / length(v) * 100},
      #extract observed integer numeric values
      #for diagnostic and exploratory purposes
      Integer_values = ~ {
        v <- suppressWarnings(as.numeric(.)) #convert to numeric
        u <- unique(v[!is.na(v) & v %% 1 == 0]) #select only non-missing integer numeric values
        #if the number of distinct values is manageable, display them explicitly
        if(length(u) <= 100) {paste(u, collapse = ", ")
        } else {"integer values > 100"}}))) %>% #otherwise avoid excessively long output
  #reshape dataset from wide to long format, obtaining one row per variable with all associated metrics
  pivot_longer(everything(),
               names_to = c("Variable", ".value"),
               names_pattern = "^(.*)_(Num_integer_values|Perc_integer_values|Integer_values)$") %>%
  #round percentage of integer values to three decimals for readability
  mutate(Perc_integer_values = round(Perc_integer_values, 3)) %>%
  #sort variables from most to least rich in integer values
  arrange(desc(Num_integer_values)) %>%
  #generate final output in HTML format
  kbl("html", caption = "Integer numeric values per variable", 
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")
```

```{r}
data_2 %>%
  select(everything()) %>% #select all variables in the dataset of interest
  #convert all variables to character to avoid implicit coercion and to manually handle numeric conversion in subsequent steps
  mutate(across(everything(), as.character)) %>%
  #compute descriptive metrics on decimal numeric values for each variable
  summarise(across(
    everything(),
    list(
      #count of decimal numeric values present in the variable
      Num_decimal_values = ~ {
        #attempt numeric conversion while suppressing warnings generated by non-convertible values
        v <- suppressWarnings(as.numeric(.))
        #count only non-missing values after conversion with a non-zero remainder in division by 1, i.e., values with a decimal component
        sum(!is.na(v) & v %% 1 != 0)},
      #compute percentage of decimal numeric values over total observations in the variable
      Perc_decimal_values = ~ {
        #convert variable to numeric
        v <- suppressWarnings(as.numeric(.))
        #compute percentage share of decimal values relative to total variable length
        sum(!is.na(v) & v %% 1 != 0) / length(v) * 100},
      #extract observed decimal numeric values for diagnostic and exploratory purposes
      Decimal_values = ~ {
        #convert to numeric
        v <- suppressWarnings(as.numeric(.))
        #select only non-missing decimal numeric values
        u <- unique(v[!is.na(v) & v %% 1 != 0])
        #if the number of distinct values is manageable, display them explicitly
        if(length(u) <= 100) {paste(u, collapse = ", ")
        } else {"decimal values > 100"}}))) %>% #otherwise avoid excessively long output
  #reshape dataset from wide to long format, obtaining one row per variable with all associated metrics
  pivot_longer(everything(),
               names_to = c("Variable", ".value"),
               names_pattern = "^(.*)_(Num_decimal_values|Perc_decimal_values|Decimal_values)$") %>%
  #round percentage of decimal values to three decimals for readability
  mutate(Perc_decimal_values = round(Perc_decimal_values, 3)) %>%
  #sort variables from most to least rich in decimal values
  arrange(desc(Num_decimal_values)) %>%
  #generate final output in HTML format
  kbl("html", caption = "Decimal numeric values per variable", 
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")
```

### **Summary statistics by variable**

```{r}
data_2 %>% 
  #base R function that returns automatic descriptive statistics for each variable:
  #for numeric variables it shows min, quartiles, median, mean, max,
  #and for categorical variables it shows frequencies or levels
  summary() %>% 
  #generate the final output in HTML format
  kbl("html", caption = "Summary statistics by variable using: summary {base}", 
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")
```

```{r}
data_2 %>% 
  select(where(is.numeric)) %>% #select only numeric variables
  psych::describe(type = 1) %>% 
  #generate the final output in HTML format
  kbl("html", caption = "Summary statistics by numeric variable using: describe {psych}", 
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")
```

```{r}
data_2 %>% 
  select(where(is.numeric)) %>% #select only numeric variables
  skimr::skim() %>% 
  #generate the final output in HTML format
  kbl("html", caption = "Summary statistics by variable using: skim {skimr}", 
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")
```

# **Associations between variables**

Because the dataset may contain both qualitative and quantitative variables (Stevens, 1946), a single correlation measure was not applied. Instead, multiple association indices were used, each appropriate for a different type of relationship between variable types. In particular, the measures used were:

1. Pearson correlation coefficient (Pearson, 1895), which measures the linear relationship between two numerical variables and is computed in R using the `cor` function from the `stats` package, specifying the "pearson" method. Its formula is:

   $$
   \rho = \frac{\sum_{i=1}^{n} (X_i - \bar{X})(Y_i - \bar{Y})}{\sqrt{\sum_{i=1}^{n} (X_i - \bar{X})^2 \sum (Y_i - \bar{Y})^2}}
   $$

   Where:

   * $X_i$ and $Y_i$ are the values of variables $X$ and $Y$,
   * $\bar{X}$ and $\bar{Y}$ are the means of variables $X$ and $Y$,
   * $\rho$ is the Pearson correlation coefficient, ranging from -1 (perfect negative correlation) to 1 (perfect positive correlation).

2. Cramér’s V (Cramér, 1946), a measure of association between two nominal variables, computed as:

   $$
   V = \sqrt{\frac{\chi^2 / n}{\min(k - 1, r - 1)}}
   $$

   Where:

   * $\chi^2$ is the chi-square test statistic for independence between the variables,
   * $n$ is the total number of observations in the dataset,
   * $k$ and $r$ are the number of categories of the two variables involved.

   This measure ranges from 0 (no association) to 1 (perfect association). In general:

   * $V > 0.1$: weak association,
   * $V > 0.3$: moderate association,
   * $V > 0.5$: strong association.

   However, in sample data, Cramér’s V tends to overestimate the strength of association. To correct this bias, the following formula is applied (Bergsma, 2013):

   $$
   \tilde{V} = \sqrt{\frac{\tilde{\varphi}^2}{\min(\tilde{k} - 1, \tilde{r} - 1)}}
   $$

   Where:

   * $\tilde{\varphi}^2 = \max\left(0, \varphi^2 - \frac{(k - 1)(r - 1)}{n - 1}\right)$ is the bias-corrected version of $\varphi^2$,
   * $\tilde{k} = k - \frac{(k - 1)^2}{n - 1}$ and $\tilde{r} = r - \frac{(r - 1)^2}{n - 1}$ are the corrections for the number of categories.

   This index was computed in R using the `cramersV` function from the `rcompanion` package (Mangiafico, 2025), applying the bias correction.

3. Spearman correlation, used to measure monotonic association (i.e., when one variable tends to increase or decrease consistently with another, even if the relationship is not linear) between two ordinal variables, or between a numerical and an ordinal variable. It is a non-parametric version of Pearson correlation, based on ranks rather than raw values. It is computed in R using the `cor` function from the `stats` package, specifying the "spearman" method. Its formula is:

   $$
   \rho_s = 1 - \frac{6 \sum d_i^2}{n(n^2 - 1)}
   $$

   Where:

   * $d_i$ is the difference between the ranks of the two variables for each observation $i$,
   * $n$ is the number of observations,
   * $\rho_s$ ranges from -1 (perfect negative monotonic correlation) to 1 (perfect positive monotonic correlation).

   Spearman correlation is particularly useful when relationships between variables are not linear but still preserve a consistent ordering.

4. Eta coefficient (Cohen, 1988), also known as the correlation ratio, is a measure used to assess the association between a categorical variable (nominal or ordinal) and a numerical variable. Unlike Pearson’s $\rho$, $\eta$ does not assume linearity in the relationship; instead, it measures the proportion of total variance explained by grouping based on the categorical variable, with values ranging from 0 (no association) to 1 (perfect association, where all variability in the numerical variable is explained by the categorical grouping).

   The general formula is:

   $$
   \eta = \sqrt{\frac{\text{BSS}}{\text{TSS}}}
   $$

   Where:

   * $\text{TSS} = \sum_{i=1}^n (Y_i - \bar{Y})^2$ is the total sum of squares around the overall mean $\bar{Y}$,
   * $\text{BSS} = \sum_{j=1}^k n_j (\bar{Y}_j - \bar{Y})^2$ is the between-group sum of squares, representing the variability explained by the categorical variable,
   * $n_j$ is the number of observations in group $j$,
   * $\bar{Y}_j$ is the mean of group $j$.

   The $\eta$ index was computed in R using a custom function that takes as input a categorical vector and a numerical vector. Categories are converted into factors to group the data, and for each group the mean and number of observations are computed, while the overall mean is calculated across all values. The total variance is then computed as the sum of squared deviations from the overall mean, and the explained variance is obtained by summing, for each group, the product of group size and squared deviation between group mean and overall mean. The ratio between explained and total variance yields $\eta^2$, whose square root gives the $\eta$ coefficient, which is finally returned as the function output.

```{r}
#mixed association matrix (Pearson, Spearman, Eta, and Cramér’s V)
mixed_assoc_matrix <- function(df) {
  #extract variable names
  vars <- names(df)
  #number of variables
  n <- length(vars)
  #initialize empty association matrix filled with NA values
  mat <- matrix(NA, n, n)
  #assign variable names to rows and columns
  colnames(mat) <- rownames(mat) <- vars
  #loop over all variable pairs
  for (i in 1:n) {
    for (j in 1:n) {
      #try to compute the association measure; if an error occurs, return NA instead of stopping the function
      mat[i, j] <- tryCatch({
        #extract the two variables
        x <- df[[i]]
        y <- df[[j]]
        #both variables are numeric: use Pearson correlation
        if (is.numeric(x) && is.numeric(y)) {
          cor(x, y, 
              method = "pearson",
              use = "pairwise.complete.obs")
          #both variables are ordinal: use Spearman correlation
          } else if (is.ordered(x) && is.ordered(y)) {
            cor(as.numeric(x), as.numeric(y),
                method = "spearman",
                use = "pairwise.complete.obs")
            #one variable is numeric and the other is ordinal: convert the ordinal variable to numeric ranks and use Spearman correlation
            } else if ((is.ordered(x) && is.numeric(y)) ||
                       (is.numeric(x) && is.ordered(y))) {
              cor(as.numeric(x), as.numeric(y),
                  method = "spearman",
                  use = "pairwise.complete.obs")
              #both variables are nominal factors: compute bias-corrected Cramér's V
              } else if (is.factor(x) && is.factor(y)) {
                cramerV(x, y, bias.correct = TRUE)
                #nominal predictor and numeric response: compute the Eta coefficient (correlation ratio)
                } else if (is.factor(x) && !is.ordered(x) && is.numeric(y)) {
                  #ensure the grouping variable is a factor
                  f <- factor(x)
                  #compute group means
                  m <- tapply(y, f, mean, na.rm = TRUE)
                  #compute group sample sizes
                  n_g <- tapply(y, f, function(z) sum(!is.na(z)))
                  #compute overall mean
                  mean_total <- mean(y, na.rm = TRUE)
                  #total sum of squares
                  TSS <- sum((y - mean_total)^2, na.rm = TRUE)
                  #between-group sum of squares
                  BSS <- sum(n_g * (m - mean_total)^2)
                  #Eta coefficient
                  sqrt(BSS / TSS)
                  #numeric predictor and nominal response: same Eta computation with variables reversed
                  } else if (is.numeric(x) && is.factor(y) && !is.ordered(y)) {
                    f <- factor(y)
                    m <- tapply(x, f, mean, na.rm = TRUE)
                    n_g <- tapply(x, f, function(z) sum(!is.na(z)))
                    mean_total <- mean(x, na.rm = TRUE)
                    TSS <- sum((x - mean_total)^2, na.rm = TRUE)
                    BSS <- sum(n_g * (m - mean_total)^2)
                    sqrt(BSS / TSS)
                    #all other variable type combinations are not supported
                    } else {NA}
        #if any computation fails (e.g., constant variable, invalid contingency table), store NA and continue with the remaining variable pairs
        }, error = function(e) NA)}}
  #round association coefficients to four decimal places
  round(mat, 4)}

rm(list = setdiff(ls(), c("data_2", "mixed_assoc_matrix"))) #clean environment removing unnecessary objects
```

```{r}
data_2_ass <- data_2 %>%
  #select variables compatible with the association measures
  #exclude unsupported data types:
  #- character variables
  #- date/time variables (Date, POSIXct, POSIXlt)
  #- time interval objects (difftime)
  #- list columns
  select(where(function(x)
    !is.character(x) &&
    !inherits(x, c("Date", "POSIXt", "difftime")) &&
    !is.list(x)))

#compute association matrix
mat_assoc <- mixed_assoc_matrix(data_2_ass)

#dataframe for HTML rendering (rounded values for display only)
mat_assoc_df <- as.data.frame(round(mat_assoc, 2))

#highlight values with different levels of association, excluding diagonal
for (i in seq_len(nrow(mat_assoc_df))) {
  for (j in seq_len(ncol(mat_assoc_df))) {
    if (i != j) {
      val <- mat_assoc[i, j] #association numeric
      #safe check for invalid values
      if (is.na(val) || is.nan(val) || is.infinite(val)) {
        mat_assoc_df[i, j] <- cell_spec("NA",color = "black",bold = TRUE)
        #perfect association (+1 or -1)
        } else if (abs(val) == 1) {
          mat_assoc_df[i, j] <- cell_spec(round(val, 2), background = "#FFD700", #gold 
                                          bold = TRUE)
          } else {
            col_fun <- grDevices::colorRampPalette(c("#B2182B", #deep red (-1)
                                                     "white",   #white (0)
                                                     "#2166AC")) #deep blue (+1)
            #map [-1, 1] -> [1, 101]
            idx <- round((val + 1) / 2 * 100) + 1
            #safety clamp for index
            idx <- max(1, min(101, idx))
            mat_assoc_df[i, j] <- cell_spec(round(val, 2), background = col_fun(101)[idx],
                                            color = ifelse(abs(val) > 0.5, "white", "black"))}}}} #text contrast

#highlight diagonal
for (i in seq_len(nrow(mat_assoc_df))) {
  mat_assoc_df[i, i] <- cell_spec(mat_assoc_df[i, i], bold = TRUE, background = "black")}

#convert to character for HTML rendering
mat_assoc_df <- mat_assoc_df %>% dplyr::mutate_all(as.character)

mat_assoc_df %>%
  #generate final output in HTML format
  kbl("html", escape = FALSE, caption = "Association matrix for mixed data", 
      align = c("l", "l")) %>% #left-align columns
  #apply HTML styling to the table
  kable_styling(bootstrap_options = c("striped",  #alternating row colors for readability
                                      "bordered", #add table borders
                                      "condensed"), #reduce spacing for a more compact table
                font_size = 10, #set table font size
                fixed_thead = TRUE) %>%  #keep header fixed while scrolling
  #make the table scrollable when it exceeds a defined height
  scroll_box(height = "400px")

rm(list = setdiff(ls(), c("data_2"))) #clean environment removing variables not needed anymore
```

# **Citations**

<!-- https://epale.ec.europa.eu/sites/default/files/apastyle2018.pdf -->
<!-- https://elearning.uniroma1.it/pluginfile.php/1085344/mod_resource/content/1/20201030_MiniguidaAPA.pdf -->

## **Bibliography**

- Bergsma, W. (2013). A bias-correction for Cramér’s V and Tschuprow’s T. *Journal of the Korean Statistical Society, 42*(3), 323–328. [https://doi.org/10.1016/j.jkss.2012.10.002](https://doi.org/10.1016/j.jkss.2012.10.002)
- Cohen, J. (1988). *Statistical power analysis for the behavioral sciences* (2nd ed.). Routledge. [https://doi.org/10.4324/9780203771587](https://doi.org/10.4324/9780203771587)
- Cramér, H. (1946). *Mathematical methods of statistics*. Princeton University Press.
- Pearson, K. (1895). Note on regression and inheritance in the case of two parents. *Proceedings of the Royal Society of London, 58*(347–352), 240–242. [https://doi.org/10.1098/rspl.1895.0041](https://doi.org/10.1098/rspl.1895.0041)
- Stevens, S. S. (1946). On the theory of scales of measurement. *Science, 103*(2684), 677–680. [https://doi.org/10.1126/science.103.2684.677](https://doi.org/10.1126/science.103.2684.677)

## **Webliography**

- Open Psychometrics Project. (n.d.). *Big Five personality test data*. Retrieved June 29, 2026, from [http://openpsychometrics.org/_rawdata/BIG5.zip](http://openpsychometrics.org/_rawdata/BIG5.zip)
- Posit Team. (2026a). *RStudio: Integrated development environment for R* (Version 2026.07.0-daily+75, “Pacific Dogwood”). Posit Software, PBC. [https://posit.co/](https://posit.co/)
- Posit Team. (2026b). *Quarto: Open-source scientific and technical publishing system* (Version 1.9.38). [https://quarto.org/](https://quarto.org/)
- R Core Team. (2026). *R: A language and environment for statistical computing* (Version 4.5.3). R Foundation for Statistical Computing. [https://www.R-project.org/](https://www.R-project.org/)

## **R Packages**

<details> 
<summary>Show complete R package citations</summary>

```{r, message=FALSE, warning=FALSE}
#| code-fold: true 
#| code-summary: "Code for R Package citations"

#This chunk extracts all currently loaded R packages (excluding base system packages) and attempts to retrieve their official citations using `citation()`. For each package, it captures the citation as text; if no citation is available, it returns a placeholder message. The results are then combined into a single formatted output, producing a readable list of package citations for the current R session.

cat(paste(lapply(
  #get all currently loaded namespaces in the session
  loadedNamespaces(),
  function(pkg) {
    #try to retrieve the package citation
    #if an error occurs, return NULL instead of stopping execution
    cit <- tryCatch(citation(pkg), error = function(e) NULL)
    #if no citation is available for the package
    txt <- if (is.null(cit)) {paste0("##", pkg, "\n", "Citation not available\n")
      } else {
        #if a citation exists:
        #- capture.output() converts citation() output into plain text
        #- paste(..., collapse = "\n") merges it into a single string
        paste0("##", pkg, "\n", paste(capture.output(cit), collapse = "\n"), "\n")}
    #return the formatted text block for this package
    txt}),
  #combine all package citations with a blank line between them
  collapse = "\n"))
```

</details> 

# <em>**Thank you for your attention from**</em> <br> <em>**Luigi Piccirillo**</em>

```{r end-time, echo=FALSE}
#print total execution time of the workflow
cat("Time required to compile the document:",
  #compute the difference between current time and stored start time in the environment
  #the result is a difftime object, which must be converted to numeric (seconds)
  round(as.numeric(Sys.time() - .time_env$start, units = "secs"), 3), "seconds")
```
